### 2015-04-24 papers/histogram/thesis/thesis.tex
updates to thesis

@@

\subsection{Introduction}
\label{sec:intro}

\fixme{Intro is confusing. Move these paragraphs in between sections
  on SW fluid and biased sampling?}

Understanding the behavior of fluids near the critical point, or the
point at which the distinction between liquid and gas ceases to become
well-defined, is a major challenge in the field of classical density
functional theory. Studying the critical point of a fluid via
simulations requires exploring the state space \fixme{remove words
  ``state space''?}  boundary between its liquid and gaseous
phases. For a generic square-well fluid to be observed in a liquid
state, the fluid must have a relatively localized distribution of the
spheres which make up the fluid. Such a distribution implies a low
energy fluid state; conversely, a low energy liquid state of a
square-well fluid otherwise observable in both liquid and gaseous
phases implies a localized distribution of spheres. The search for
liquid states is thus equivalent to the search for low energy
states. In unbiased Monte Carlo simulations, however, highly localized
distributions of spheres are extremely unlikely to occur, making such
simulations impractical for studying the square-well fluid.

Several generic computational methods exist for dealing with such a
problem; namely, that interesting regions of a system's state space
cannot be sampled via unbiased Monte Carlo simulations in a reasonable
amount of time. There is no consensus, however, on the relative
efficacy of these methods. Furthermore, some methods, such as the
Wang-Landau method\cite{wang_landau}, require tuning several free
parameters that lack both ``obvious'' values and systematic means for
assigning them. This work will thus attempt to quantify and compare
the effectiveness of some such methods.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]{The square-well pair
    potential can be used to model short range forces to first
    order. The infinite potential at $r<\sigma$ simply enforces the
    condition that spheres cannot overlap. At $r>\lambda\sigma$, the
    potential is zero, and makes no contribution to the fluid's
    internal energy.}
  \label{fig:pair_potential}
\end{figure}

The square-well (SW) fluid is a simple model used in classical density
functional theories to capture low order effects of short-range
attractive forces, such as the van der Waals force. The fluid is
composed of spheres with diameter $\sigma$ which have a pair potential
\begin{align}
  v_{sw}\p{\vec r}=\left\{
    \begin{array}{ll}
      \infty & \abs{\vec r}<\sigma \\
      -\epsilon & \sigma<\abs{\vec r}<\lambda\sigma \\
      0 & \abs{\vec r}>\lambda\sigma
    \end{array}
  \right., \label{eq:pair_potential}
\end{align}
also shown graphically in Figure \ref{fig:pair_potential}, where the
parameters $\lambda$ and $\epsilon$ are referred to as the well width
and depth, respectively. The first ($\abs{\vec r}<\sigma$) part of
this potential forbids spheres from overlapping, whereas the second
($\sigma<\abs{\vec r}<\lambda\sigma$) associates an energy $-\epsilon$
with each pair of spheres whose centers are within distance of
$\lambda\sigma$ of each other (where typically
$\lambda\in\sp{1.05,3}$). The net potential energy of the square-well
fluid is thus
\begin{align}
  E=\sum_{i<j}v_{sw}\p{\vec r_i-\vec r_j},
  \label{eq:internal_energy}
\end{align}
where $\vec r_i$ is the position of the $i$-th sphere. As the
potential energy $E$ is the primary form of energy concerning us in
this paper, we will refer to $E$ as simply the ``energy'' of the
fluid. An important feature of the square-well fluid is that its
energy is always an integer multiple of the well depth. A homogeneous
square-well fluid is thus uniquely identified by its well width
$\lambda$ and filling fraction $\eta$ (i.e. the proportion of space
filled by spheres; a dimensionless density), as all other properties
can be normalized to the natural energy scale $\epsilon$ and length
scale $\sigma$. In practice, our simulation codes use dimensionless
energies $E/\epsilon$, temperatures $kT/\epsilon$, and distances
$r/\sigma$.

\subsection{Monte Carlo fluid simulations}
\label{sec:MC_simulations}

While model systems are powerful tools for understanding complex
physical systems, they do not themselves exist in the real
world. Consequently, direct experimental tests of theories for model
system (e.g. square-well density functional theories) are not
possible. For this reason, model system theories are commonly tested
against Monte Carlo simulations. Proper implementation of Monte Carlo
methods to study completely characterized systems ensures that
statistical results from simulations converge on the exact properties
of the simulated system in the infinite simulation time
limit\fixme{explain. cite?}. Furthermore, uncertainties in quantities
computed via Monte Carlo simulations are typically well-defined,
monotonically decreasing functions of simulation time, allowing one to
run simulations to the desired level of accuracy.

\begin{algorithm}[!b]
  \caption{Unbiased Monte Carlo fluid simulation}
  \label{alg:MC}
  \begin{alg}

@@

  \item Attempt to change the position of one sphere (in general, a
    single fluid ``atom'' or ``molecule'') to a random location,
    rejecting the change if it results in a forbidden fluid
    configuration (e.g. two or more spheres overlap) and accepting the
    change otherwise. Attempting to move a single sphere is referred
    to as a move. \label{alg:MC_move}

  \item Repeat step \ref{alg:MC_move} for every other sphere in the
    fluid. Attempting to move every sphere once is referred to as an
    iteration of the simulation.
    \label{alg:MC_iteration}

  \item Repeat step \ref{alg:MC_iteration} indefinitely, or until data
    of sufficient quality has been generated, periodically collecting
    and dumping statistics on fluid states (e.g. energy, pair
    distribution histograms, etc.) to data files.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased Monte Carlo fluid
simulations. Such an algorithm is ``unbiased'' in the sense that it
collects statistics (i.e. data) on all valid system configurations
with equal probability. Statistics whose collection time scales as
$\mathcal O\p{1}$ with system size can be collected after every move
(defined in the algorithm), while statistics whose collection time
scales as $\mathcal O\p{N}$, where $N$ is the number of simulated
elements (e.g. spheres), can be collected after each iteration. In
general, collection with $\mathcal O\p{\chi}$ time scaling in $N$
should not occur more often than once every $\chi$ moves
\fixme{explain better?}, so as to ensure that scaling up simulations
does not cause them to asymptotically spend all of their time
collecting statistics, and no time actually simulating the
fluid. Collected statistics are used to find thermodynamic properties
of the simulated fluid.

In this work, we are concerned with simulating the {\emph homogeneous}
square-well fluid. To avoid edge effects resulting from fluid behavior
near a wall, we employ periodic boundary conditions. The use of a
finite cell with periodic boundary conditions suppresses all density
fluctuations on scales larger than the dimensions of the simulated
fluid cell, thereby introducing a source of error; addressing this
error, however, is outside the scope of this work. \fixme{Add
  feedback}

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ (e.g. energy) \fixme{other
  examples} is directly proportional to the density of states $D\p{X}$
in $X$ \fixme{explain?}. It is sometimes the case, however, that the
density of states in some region $R$ in the range of possible $X$ is
so low that it is practically impossible to sufficiently sample $R$,
that is, accumulate a statistically significant $H\p{R}$, in any
reasonable amount of time.

@@

Histogram methods provide a means to address unbiased Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:MC_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

\begin{algorithm}
  \caption{Biased Monte Carlo fluid simulation}
  \label{alg:biased_MC}
  \begin{alg}

@@

  \item Attempt to change the position of one sphere, rejecting the
    change if
    \begin{enumerate*}[label=\roman*)]
    \item it results in a forbidden fluid configuration, or
    \item a newly chosen random number on the interval $\sp{0,1}$ is
      larger than the probability given by (\ref{eq:move_prob}).
    \end{enumerate*}
    \label{alg:biased_MC_move}

  \item Repeat step \ref{alg:biased_MC_move} for every other sphere in
    the fluid.
    \label{alg:biased_MC_iteration}

  \item Repeat step \ref{alg:biased_MC_iteration} until data of
    sufficient quality has been generated.

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1. If we wish to
convert a histogram $H\p{X}$ of observations in a biased Monte Carlo
simulation into a numerical unnormalized density of states $\tilde
D\p{X}$ in $X$, we therefore divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos_unnormed}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A ``histogram method'' is an algorithm for determining a
weight function appropriate for a particular simulation.

@@

@@

\fixme{Put canonical before this?}

The sort of Monte Carlo simulations described in this paper are only
capable of collecting statistics on system microstates, and functions
of microstates. For example, in simulation one might periodically
compute both the energy $E\p{s}$ and some system property $X\p{s}$,
both of which are determined by the system microstate $s$, in order to
find the mean value of $X$ at any given energy $E$,
i.e. $\bk{X}_E$. In the real world, however, information about a
system's microstate and functions of microstates (e.g. the energy $E$)
is generally is inaccessible. One therefore cannot measure $\bk{X}_E$
directly. Instead, one typically measures the dependence of
thermodynamic properties on macroscopic state variables such as
temperature, i.e. $\bk{X}_T$.

The temperature dependence of $\bk{X}_T$ can be expressed in terms of
the density of states $D\p{E}$ and the expectation value $\bk{X}_E$ by
\begin{align}
  \bk{X}_T=\f1{Z\p{T}}\sum_E\bk{X}_ED\p{E}e^{-E/kT},
  \label{eq:temp_dep_X_norm}
\end{align}
where the partition function $Z\p{T}$ is simply a normalization
factor, given by
\begin{align}
  Z\p{T}=\sum_ED\p{E}e^{-E/kT}.
\end{align}
To reduce redundant computations and numerical error in
implementation, we will use a partition function with an unnormalized
density of states,
\begin{align}
  \tilde Z\p{T}=\sum_E\tilde D\p{E}e^{-E/kT},
\end{align}
in terms of which
\begin{align}
  \bk{X}_T=\f1{\tilde Z\p{T}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT}.
  \label{eq:temp_dep_X}
\end{align}
Sufficient accumulation of statistics on $\bk{X}_E$ at all available
energies thus in principle allows one to determine $\bk{X}_T$ for any
$T$. In practice, the density of states can fall off so quickly with
energy that some range of allowable energies is practically
inaccessible via Monte Carlo simulations, biased or otherwise. In such
a case, computing $\bk{X}_T$ to a reasonable degree of accuracy
requires sufficiently sampling the energies at which $\tilde
D\p{E}e^{-E/kT}$ dominates the sum in (\ref{eq:temp_dep_X}).

We will employ histogram methods in order to sample as broad of an
energy range as possible in each simulation. Broad energy sampling
will allow us to determine, for various square-well fluids,
\begin{enumerate*}[label=\roman*)]
\item the density of states (over energy), and
\item the temperature dependence of various thermodynamic properties,
  particularly at temperatures near the liquid-gas phase boundary.
\end{enumerate*}

@@

\subsection{Canonical (fixed temperature) weights}
\label{sec:canonical_weights}

A standard construction of the weight array $w\p{E}$ for what is
called a canonical Monte Carlo simulation involves \fixme{better word}
choosing a particular temperature $T_0$, and using weights
proportional to the Boltzmann factor at that temperature, i.e.
\begin{align}
  w\p{E}=e^{-E/kT_0},
\end{align}
where there is no reason to normalize $w\p{E}$ due to the fact that
only ratios of weights, as per (\ref{eq:move_prob}), are ever used in
simulation. The fact that only ratios of weights are used in
simulation makes $w\p{E}$ scale-invariant.

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT}
  =\sum_EH\p{E}/w\p{E}~e^{-E/kT} =\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $X\p{T_0}$
\begin{align}
  X\p{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT_0}
  =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}}.
  \label{eq:X_canonical}
\end{align}
The simplifications in in (\ref{Z_canonical}) and
(\ref{eq:X_canonical}), which have no explicit dependence on $T_0$,
occur because canonical Monte Carlo simulations sample energies in
proportion to the distribution (over energy) of microstates at a
temperature of $T_0$.  Canonical Monte Carlo simulations can therefore
fail to sufficiently sample energies which are important
(i.e. energies with a non-negligible state probability density) at
different temperatures. As a consequence, such simulations should not
be used to determine properties $X\p{T\ne T_0}$, and are thus referred
to as ``fixed temperature'' simulations.

Due to these problems \fixme{be more explicit?}, we will not use
canonical weights alone to study the square-well fluid. We will,
however, use canonical weights for part of all weight arrays, as
discussed in Section \ref{sec:min_energy}.

@@

The multi-canonical histogram method assumes complete knowledge of the
density of states $D\p E$ of the system in question, and solves
(\ref{eq:dos_unnormed}) for a weight array $w\p{E}$ which should yield
a flat energy histogram $H\p E=H_0$ to get
\begin{align}
  w\p E=\f1{D\p E},
  \label{eq:flat_weights}
\end{align}
where we may neglect the constant scale factor $H_0$ and do not worry
about normalization factor distinguishing $D\p{E}$ from $\tilde
D\p{E}$ due to the scale-invariance of $w\p{E}$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

\fixme{talk more about phase transitions and the critical point?}

Understanding the behavior of fluids near the critical point, or the
point at which the distinction between liquid and gas ceases to become
well-defined, is a major challenge in the field of classical density
functional theory. Studying the critical point of a fluid via
simulations requires exploring the boundary between its liquid and
gaseous phases. The square-well fluid (Section \ref{sec:sw_fluid}) is
the simplest system with a liquid-vapor phase transition, and is
therefore of interest for studying the critical point of fluids. For a
generic square-well fluid to be observed in a liquid state, it must
have a localized distribution of the spheres which make up the
fluid. Such a distribution implies a low energy fluid state, as all
interactions between spheres are attractive. Conversely, a low energy
liquid state of a square-well fluid otherwise observable in both
liquid and gaseous phases implies a localized distribution of
spheres. The search for liquid states of the square-well fluid is thus
equivalent to the search for low energy states.

Monte Carlo simulations (Section \ref{sec:mc_sim}) are a standard
method for studying the thermodynamic properties of a system. In the
most straightforward implementation of Monte Carlo simulations,
however, highly localized distributions of spheres are extremely
unlikely to occur, making such simulations impractical for studying
the critical point of the square-well fluid. Several generic
computational methods, called histogram methods (Section
\ref{sec:histogram_methods}), exist for dealing with such a problem;
namely, that interesting regions of a system's state space cannot be
sampled via standard Monte Carlo simulations in a reasonable amount of
time. There is no consensus, however, on the relative efficacy of
these methods. Furthermore, some methods, such as the Wang-Landau
method\cite{wang_landau}, require tuning several free parameters that
lack both ``obvious'' values and systematic means for assigning
them. This work will thus attempt to quantify and compare the
effectiveness of a few such methods.

@@

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]{The square-well pair
    potential can be used to model short range forces to first
    order. The infinite potential at $r<\sigma$ simply enforces the
    condition that spheres cannot overlap. At $r>\lambda\sigma$, the
    potential is zero, and makes no contribution to the fluid's
    internal energy.}
  \label{fig:pair_potential}
\end{figure}

The square-well (SW) fluid is a simple model used in classical density
functional theories to capture low order effects of short-range
attractive forces, such as the van der Waals force. The fluid is
composed of spheres with diameter $\sigma$ which have a pair potential
\begin{align}
  v_{sw}\p{\vec r}=\left\{
    \begin{array}{ll}
      \infty & \abs{\vec r}<\sigma \\
      -\epsilon & \sigma<\abs{\vec r}<\lambda\sigma \\
      0 & \abs{\vec r}>\lambda\sigma
    \end{array}
  \right., \label{eq:pair_potential}
\end{align}
also shown graphically in Figure \ref{fig:pair_potential}, where the
parameters $\lambda$ and $\epsilon$ are referred to as the well width
and depth, respectively. The first ($\abs{\vec r}<\sigma$) part of
this potential forbids spheres from overlapping, whereas the second
($\sigma<\abs{\vec r}<\lambda\sigma$) associates an energy $-\epsilon$
with each pair of spheres whose centers are within distance of
$\lambda\sigma$ of each other (where typically, $\lambda\in(1,3]$).
The net potential energy of the square-well fluid is thus
\begin{align}
  E=\sum_{i<j}v_{sw}\p{\vec r_i-\vec r_j},
  \label{eq:internal_energy}
\end{align}
where $\vec r_i$ is the position of the $i$-th sphere. As the
potential energy $E$ is the primary form of energy concerning us in
this paper, we will refer to $E$ as simply the ``energy'' of the
fluid. An important feature of the square-well fluid is that its
energy is always an integer multiple of the well depth. A homogeneous
square-well fluid is thus uniquely identified by its well width
$\lambda$ and filling fraction $\eta$ (i.e. the proportion of space
filled by spheres; a dimensionless density), as all other properties
can be normalized to the natural energy scale $\epsilon$ and length
scale $\sigma$. In practice, our simulation codes use dimensionless
energies $E/\epsilon$, temperatures $kT/\epsilon$, and distances
$r/\sigma$.

\subsection{Monte Carlo fluid simulations}
\label{sec:mc_sim}

While model systems are powerful tools for understanding complex
physical systems, they do not themselves exist in the real
world. Consequently, direct experimental tests of theories for model
system (e.g. square-well density functional theories) are not
possible. For this reason, model system theories are commonly tested
against Monte Carlo simulations. Proper implementation of Monte Carlo
methods to study completely characterized systems ensures that
statistical results from simulations converge on the exact properties
of the simulated system in the infinite simulation time
limit. Furthermore, uncertainties in quantities computed via Monte
Carlo simulations are typically well-defined, monotonically decreasing
functions of simulation time, allowing one to run simulations to the
desired level of accuracy.

\subsubsection{Implementation}
\label{sec:mc_implementation}

\begin{algorithm}[tb]
  \caption{Unbiased Monte Carlo fluid simulation}
  \label{alg:MC}
  \begin{alg}

@@

  \item Attempt to change the position of one sphere (in general, a
    single fluid ``atom'' or ``molecule'') to a random location,
    rejecting the change if it results in a forbidden fluid
    configuration (e.g. two or more spheres overlap) and accepting the
    change otherwise. Attempting to move a single sphere is referred
    to as a move. \label{alg:mc_move}

  \item Repeat step \ref{alg:mc_move} for every other sphere in the
    fluid. Attempting to move every sphere once is referred to as an
    iteration of the simulation.
    \label{alg:mc_iteration}

  \item Repeat step \ref{alg:mc_iteration} indefinitely, or until data
    of sufficient quality has been generated, periodically collecting
    and dumping statistics on fluid states (e.g. energy, pair
    distribution histograms, etc.) to data files.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased Monte Carlo fluid
simulations. Such an algorithm is ``unbiased'' in the sense that it
collects statistics (i.e. data) on all valid system configurations
with equal probability. Statistics whose collection time scales as
$\mathcal O\p{1}$ with system size, meaning that increasing the number
of spheres $N$ in the simulation does not affect collection time, can
be collected after every move (defined in the algorithm). Statistics
whose collection time scales as $\mathcal O\p{N}$, meaning that
doubling $N$ doubles collection time, can be collected after each
iteration. In general, collection with $\mathcal O\sp{\chi\p{N}}$ time
scaling should not occur more often than once every $\chi\p{N}$ moves
(where $\chi\p{N}$ may be an arbitrary function of only $N$ void of
constant prefactors, e.g. $N\log N$, or $2^N$). These collection rules
ensure that scaling up simulations does not cause them to
asymptotically spend all of their time collecting statistics, and no
time actually simulating the fluid. Collected statistics are used to
find thermodynamic properties of the simulated fluid.

In this work, we are concerned with simulating the \emph{homogeneous}
square-well fluid. To avoid edge effects resulting from fluid behavior
near a wall, we employ periodic boundary conditions. The use of a
finite cell with periodic boundary conditions suppresses all density
fluctuations on scales larger than the dimensions of the simulated
fluid cell, thereby introducing a source of error. Addressing and
sequestering this error, however, is outside the scope of this work,
and involves considering the limit of numerical results as the number
of spheres $N\to\infty$ (with a constant filling fraction).

\subsubsection{Computing observables}
\label{sec:computing_observables}

The sort of Monte Carlo simulations described above are only capable
of collecting statistics on functions of system microstates, and
finding correlations between these functions. For example, in
simulation one might periodically compute both the energy $E\p{s}$ and
some system property $X\p{s}$, both of which are determined by the
microstate $s$, in order to find the mean value of $X$ at any given
energy $E$, i.e. $\bk{X}_E$. In the real world, however, information
about a system's microstates, and by extension functions of
microstates (e.g. the energy $E$), is inaccessible. One therefore
cannot measure $\bk{X}_E$ directly. Instead, one typically measures
the dependence of thermodynamic properties on macroscopic state
variables such as temperature, i.e. $\bk{X}_T$.

To find $\bk{X}_T$, we first need to understand the concept of a
density of states. Given an arbitrary system property $Y\p{s}$
(e.g. energy) determined by the microstate $s$, one can define a
density of states $D\p{Y}$ such that
\begin{enumerate*}[label=\roman*)]
\item $D\p{Y_0}$ is proportional to the number of microstates $s$ for
  which $Y\p{s}=Y_0$, and
\item $\sum_YD\p{Y}=1$.
\end{enumerate*}
The temperature dependence of $\bk{X}_T$ can be expressed in terms of
the expectation value $\bk{X}_E$ and the density of states $D\p{E}$ by
\begin{align}
  \bk{X}_T=\f1{Z\p{T}}\sum_E\bk{X}_ED\p{E}e^{-E/kT},
  \label{eq:XT_norm}
\end{align}
where the partition function $Z\p{T}$ is simply a normalization
factor, given by
\begin{align}
  Z\p{T}=\sum_ED\p{E}e^{-E/kT}.
\end{align}
To reduce redundant computations and numerical error in
implementation, we will use a partition function with an unnormalized
density of states,
\begin{align}
  \tilde Z\p{T}=\sum_E\tilde D\p{E}e^{-E/kT},
\end{align}
in terms of which
\begin{align}
  \bk{X}_T=\f1{\tilde Z\p{T}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT}.
  \label{eq:XT}
\end{align}

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
unbiased Monte Carlo) to sufficiently sample $R$, that is, accumulate
a statistically significant histogram $H\p{R}$, in any reasonable
amount of time.

@@

Histogram methods provide a means to address unbiased Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:mc_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

\begin{algorithm}[tb]
  \caption{Biased Monte Carlo fluid simulation}
  \label{alg:biased_MC}
  \begin{alg}

@@

  \item Attempt to change the position of one sphere, rejecting the
    change if
    \begin{enumerate*}[label=\roman*)]
    \item it results in a forbidden fluid configuration, or
    \item a newly chosen random number on the interval $\sp{0,1}$ is
      larger than the probability given by (\ref{eq:move_prob}).
    \end{enumerate*}
    \label{alg:biased_mc_move}

  \item Repeat step \ref{alg:biased_mc_move} for every other sphere in
    the fluid.
    \label{alg:biased_mc_iteration}

  \item Repeat step \ref{alg:biased_mc_iteration} until data of
    sufficient quality has been generated.

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1. If we wish to
convert a histogram $H\p{X}$ of observations in a biased Monte Carlo
simulation into a numerical unnormalized density of states $\tilde
D\p{X}$ in $X$, we therefore divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos_norm}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A ``histogram method'' is an algorithm for determining a
weight function appropriate for a particular simulation.

@@

\subsubsection{Canonical weights}
\label{sec:canonical_weights}

The most common weight array $w\p{E}$ used by physicists for what is
called a canonical Monte Carlo simulation involves choosing a
particular temperature $T_0$, and using weights proportional to the
Boltzmann factor at that temperature, i.e.
\begin{align}
  w\p{E}=e^{-E/kT_0},
\end{align}
where there is no reason to normalize $w\p{E}$ due to the fact that
only ratios of weights, as per (\ref{eq:move_prob}), are ever used in
simulation. The fact that only ratios of weights are used in
simulation makes $w\p{E}$ scale-invariant.

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{w\p{E}}~e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{e^{-E/kT_0}}~e^{-E/kT_0}=\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $X\p{T_0}$
\begin{align}
  X\p{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT_0}
  =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}}.
  \label{eq:X_canonical}
\end{align}
The simplifications in in (\ref{Z_canonical}) and
(\ref{eq:X_canonical}), which have no explicit dependence on $T_0$,
occur because canonical Monte Carlo simulations sample energies in
proportion to the distribution (over energy) of microstates at a
temperature of $T_0$.  Canonical Monte Carlo simulations can therefore
fail to sufficiently sample energies which are important
(i.e. energies with a non-negligible state probability density) at
different temperatures. As a consequence, such simulations should not
be used to determine properties $X\p{T\ne T_0}$, and are thus referred
to as ``fixed temperature'' simulations.

Due to these problems, we will not use canonical weights alone to
study the square-well fluid. We will, however, use canonical weights
for part of all weight arrays, as discussed in Section
\ref{sec:min_energy}.

@@

Given the expression for $\bk{X}_T$ in (\ref{eq:XT}), sufficient
accumulation of statistics on $\bk{X}_E$ at all available energies in
principle allows one to determine $\bk{X}_T$ for any temperature
$T$. In practice, the density of states can fall off so quickly with
energy that some range of allowable energies is practically
inaccessible via Monte Carlo simulations, biased or otherwise. In such
a case, computing $\bk{X}_T$ to a reasonable degree of accuracy
requires sufficiently sampling the energies at which $\tilde
D\p{E}e^{-E/kT}$ dominates the sum in (\ref{eq:XT}).

We will employ histogram methods in order to sample as broad of an
energy range as possible in each simulation. Broad energy sampling
will allow us to determine, for various square-well fluids,
\begin{enumerate*}[label=\roman*)]
\item the density of states $D\p{E}$, and
\item the temperature dependence $\bk{X}_T$ of various thermodynamic
  properties $X$, particularly at temperatures near the liquid-gas
  phase boundary.
\end{enumerate*}

@@

@@

The multi-canonical histogram method assumes complete knowledge of the
density of states $D\p E$ of the system in question, and solves
(\ref{eq:dos}) for a weight array $w\p{E}$ which should yield a flat
energy histogram $H\p E=H_0$ to get
\begin{align}
  w\p E=\f1{D\p E},
  \label{eq:flat_weights}
\end{align}
where we may neglect the constant scale factor $H_0$ and do not worry
about normalization factor distinguishing $D\p{E}$ from $\tilde
D\p{E}$ due to the scale-invariance of $w\p{E}$.
